A primer to numerical simulations: The perihelion motion of Mercury

### 2018-01-15 tex/MercuryNumerical.tex
Updated simulation values
Update the simulation parameters and interpolation range in order to get closer to the numerical result by GR. I have al
so updated the example script, plots and added a short description of what I have been doing. This must still be complet
ed. Furthermore, the publication scripts must be updated and a summary must be written.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example: General Relativity and the perihelion motion of Mercury}\label{sec:gr}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For this concrete example the starting point for the force is Newton's law of gravitation
\begin{equation}
\vec F_N(\vec r) = - \frac{G_N m M_\odot}{r^2} \frac{\vec r}{r}\ ,
\end{equation}
where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation, $m$ is the mass of Mercury and $M_\odot=2\times 10^{30}$ kg is the mass of the Sun.
In addition $r=|\vec r(t)|$ denotes the distance between Sun and Mercury when we assume that the Sun is infinitely heavier than Mercury and located at the center of the coordinate system.
Although this is not exact, it is a pretty good approximation because $m/M_\odot\sim 10^{-8}$.
For later convenience we introduce the Schwarzschild radius of the Sun
\begin{equation}
r_S=\frac{2G_N  M_\odot}{c^2} = 3 \ \mbox{km} \ , \label{rsdef}
\end{equation}
where $c=3\times 10^8$ m/s denotes the speed of light.
Note that $r_S$ is the characteristic length scale of the gravitational field of the Sun for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and $c$.
The Schwarzschild radius is also an important quantity to characterize black holes; this is however not relevant for the discussion at hand.
With this, Newton's second law reads
\begin{equation}
\ddot{\vec r}      = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\frac{\vec r}{r} \, . \label{eq:newton}
\end{equation}
An attractive force that vanishes at large distances leads in general, depending on the initial conditions, either to bounded or to open orbits.
In the latter case the planet simply disappears from the Sun, since a given force can capture only bodies with small enough momenta.
The students may study those scenarios within their simulations by varying the start velocity while keeping the start location fixed.

@@

To allow for a movement of the perihelion we need to modify the force that led to Eq.~(\ref{eq:newton}).
This modification should be such that it depends on $r$ and it vanishes for $r\to \infty$.
Based on the discussion presented above we may multiply the right hand side of~(\ref{eq:newton}) with a factor
$(1+\alpha(r_S/r))$, where $\alpha$ denotes some dimensionless parameter.
This seems like a natural way to parametrize the additional potential because it adds the smallest possible deviation from an $1/r$ potential expressed relative to the characteristic length $r_S$.
However, besides $r_S$, which characterizes the gravitational potential of the Sun, there is also a characteristic parameter for the dynamics of Mercury:
\begin{equation}
r_a^2 := \frac{\vec L\,^2}{m^2c^2}= \frac{{(\vec r\times \dot{\vec r} \, )}^2}{c^2} \ , \label{a2def}
\end{equation}
where $\vec L$ denotes the angular momentum, which is a constant of motion for central potentials.
It is easily verified that $r_a^2$ carries dimensions of length squared.
We therefore have to add one more term to the potential; in analogy to the one discussed above, we use $\beta (r_a^2/r^2)$.
Combining everything, we use the following ansatz for the modified equation of motion:
\begin{equation}
\ddot{\vec r} = - \frac{c^2}{2}\frac{r_S}{r^2}\left(1+\alpha\frac{r_S}{r}+\beta\frac{r_a^2}{r^2}\right) \, \frac{\vec{r}}{r} \ .
\label{eq:newton_art}
\end{equation}
For the system at hand, $r_a^2$ may be estimated from the parameters of Mercury at its perihelion: The corresponding velocity is
 $\dot{\vec r}(t=0)=\abs{\vec v_M(0)} = 59$~km/s (here we already indicate that we will start the simulation at the perihelion) and the closest distance between Sun and Mercury is
$ r_{MS}=\abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km~\cite{MercuryFactSheet}. Note that at the perihelion $\vec r$ and $\dot{\vec r}$ are perpendicular to each other.
We thus find that $(r_a^2/ r_{MS}^2) = 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})=6\cdot 10^{-8}$.
Therefore (for a more detailed discussion about the underlying logic we refer to Sec.~\ref{sec:analysis}) one may expect from both correction terms an effect of similar size. Furthermore, any term of higher order in either $a$ or $r_S$
should be suppressed by additional seven orders of magnitude and thus can be neglected safely.
The parameters $\alpha$ and $\beta$ can be extracted from a fit to data --- this will be done below.
However, since we have only a single number to study, the perihelion shift (see Eq.~(\ref{delT})),
we can only fix either $\alpha$ or $\beta$ (or a linear combination thereof).
The parameters can also be calculated from the underlying theory one uses. The actual values depend on the specific theory;
General relativity gives~\cite{Einstein}
\begin{equation}
	\alpha = 0 \, , \quad \beta = 3 \, . \label{eq:art-prediction}
\end{equation}
Thus we may also use the simulation explained below to calculate the perihelion motion of Mercury from the input
values given in Eq.~(\ref{eq:art-prediction}).

@@

Since the trajectories do not depend on where on the orbit of Mercury the simulation is started, we
use the values at the perihelion with $\abs{\vec r_{MS}(0)} = 46 \cdot 10^6\,\mathrm{km}$ and $\abs{\vec v_M(0)} = 59\,\mathrm{km/s}$\cite{MercuryFactSheet} as initial (for $t=0$) parameters.
The computer does not understand physical units, hence one has to express each variable in an appropriate unit.
Both for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range", e.g.,
by expressing distances in $R_0 = 10^{10}\,\mathrm{m}$ and time intervals in days, $T_0 = 1\,\mathrm{d}$, where d refers to earth days. One Mercury year is given by $T_M=88\,T_0$.
With this choice, the initial distance of Mercury to the Sun, the size of the initial velocity of Mercury and the acceleration prefactor (see Eq.~(\ref{eq:newton_art})) become
\begin{eqnarray}
r_{MS}(0)   &=& 4.6 R_0 \, , \quad
v_{M}(0)    = 0.51 \frac{R_0}{T_0} \, ,  \\
a_{MS}(r_{MS}) &=& \frac{c^2}{2}\frac{r_S}{r^2} = 1.01 \frac{R_0}{T_0^2} \frac{1}{{\left(r_{MS}/R_0\right)}^2}\, .
\end{eqnarray}

@@

In \python{}, this reads
\begin{lstlisting}
# Definition of parameters
rM0 = 4.6    # in units of R0
vM0 = 0.51   # in units of R0/T0
c_a = 1.01   # in units of R0**3/T0**2
TM  = 88.    # in units of T0
rS  = 3.0e-7 # in units of R0
ra2 = 8.2e-7 # in units of R0**2
\end{lstlisting}
where $\texttt{c\_a} =a_{MS} r_{MS}^2$, the acceleration without the radius factor.
The last two quantities refer to the Schwarzschild radius of the Sun and the parameter $r_a^2$ defined in Eqs. (\ref{rsdef}) and (\ref{a2def}), respectively.

@@

If the starting values are chosen as advised in the previous section, the students should end up with a trajectory as depicted in figure~\ref{fig:MercuryOrbit-a0-small}.
At this point one might ask students to vary $\Delta t$ and observe the effect of this on the orbit [see also figure~\ref{fig:MercuryOrbit-a0-small-dt-large}].
To get the perihelion motion, the additional force term described in Eq.~(\ref{eq:newton_art}) has to be "turned on", e.g., by choosing  $\alpha\neq0$
[figure~\ref{fig:MercuryOrbit-a6-small}].
This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for its impact on the trajectories. In particular
they should discover that values of $\alpha$ of the order of $10^6$ are necessary to get a visible effect.
For a nicely visible perihelion motion it is advisable to not use the correct Mercury values but a more excentric trajectory by choosing, e.g., $r_{MS}(0)=6R_0$ [figure~\ref{fig:MercuryOrbit-a6-big}].
\begin{figure}[htb]
	\centering
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a0T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a0-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/num-err.png}
		\caption{\label{fig:MercuryOrbit-a0-small-dt-large}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T10dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-big}}
	\end{subfigure}
	\captionsetup{singlelinecheck=off}
	\caption[]{\label{fig:MercuryOrbit}
		Different Mercury orbits for $\beta=0$ and
		\begin{itemize}
		\item[(a)] $\alpha=0$ and $\Delta t = \Delta t_0  / 20$;
		\item[(b)] $\alpha=0$ and $\Delta t = \Delta t_0  \times 2$, where the black line represents the orbit of (a) and the red line is the orbit for the larger time steps;
		\item[(c)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$;
		\item[(d)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$ but $r_{MS}(0) = 6 R_0$.
		\end{itemize}
		Here, the time steps are defined by $\Delta t_0 \equiv 2 v_{M}(0) / c_a$ and the images are screenshots of the simulation (with modified colors).
		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "Mercury year" changes.
	}
\end{figure}

@@

\begin{figure}[htb]
	\centering
	\begin{subfigure}[c]{0.49\textwidth}
		\includegraphics[width=\textwidth]{figs/alpha-angle.pdf}
		\caption{\label{fig:angle-alpha}}
	\end{subfigure}
	\begin{subfigure}[c]{0.49\textwidth}
		\includegraphics[width=\textwidth]{figs/beta-angle.pdf}
		\caption{\label{fig:beta-alpha}}
	\end{subfigure}
	\caption{\label{fig:AlphaAngle} Linear relation between $\alpha$ (a), $\beta$ (b) and the perihelion motion $\delta \Theta$ for $\Delta t=2v_M(0)/c_a/100$ and the other parameter set to zero.}
\end{figure}

As an example we  extract the perihelion motion using the parameters provided by General Relativity, namely $\alpha = 0$ and $\beta = 3$.
Considering one result from the numeric simulation, e.g., $\delta\Theta(\alpha = 0, \beta=10^6) \approx 10.7^{\degree}$, the sought-after angle can be calculated by a linear interpolation as
\begin{equation}
	\delta\Theta (\alpha=0, \beta=3) = \frac{10.7^{\degree}}{10^6} \times 3 = 0.116''
	\, .
\end{equation}
Depending on the knowledge of the students, multiple points and also estimated numerical uncertainties could be used to extract this value by a linear regression.
Note that, when comparing the results of the simulation with the experimental result, one Earth year corresponds to $T_E/T_M \approx 365d/88d\approx 4.15$ Mercury years.
Therefore, the perihelion motion per 100 earth years is $0.116''\cdot 415\approx 48.2''$.
This is close to the experimental value of $\delta\Theta_{\mathrm{exp}} = 43.0''$.

@@

\begin{figure}[htb]
	\centering
	\includegraphics[width=.99\textwidth]{figs/angular-variaton.pdf}
	\caption{\label{fcc3}Motion of perihelion $\delta\Theta$ in degrees depending on the number of turns $N_t$ for different time steps $\Delta t$ (color) and different values for $\beta$ (columns).
	Here, $\Delta t_0 \equiv 2 v_M(0)/c_a$.
	Offsets of $\delta \Theta(\alpha=0, \beta=0)$ from zero can be used as an estimated of the magnitude of the error. Oscillations in $\delta \Theta$ indicate too coarse time steps.
}
\end{figure}

@@

This problem can be pointed out to the students by asking them to vary the size of the time-steps and observe the effect on the trajectories.
This should also be done for a wide range of values to show both good and problematic regimes.
It is advantageous to set both $\alpha$ and $\beta$ to zero for this study, since in this case the trajectories should be closed and deviations from the correct case can be spotted easily.
Figure~\ref{fig:MercuryOrbit-a0-small-dt-large} shows the Mercury trajectory for $\Delta t=2 v_M(0) / c_a \times 2$.
One can clearly see the failure of the simulation to reproduce the physical trajectory.
From examples like this it should be apparent that the time steps influence the accuracy of the simulation.
The simulation reproduces the actual trajectory only in the limit $\Delta t \rightarrow 0$. Thus, in any numerical simulation
one always has to identify a proper compromise between numerical accuracy and time spent for the simulation
(clearly for $\Delta t\to 0$ the computing time goes to infinity).

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example: General Relativity and the perihelion motion of Mercury}\label{sec:gr}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For this concrete example the starting point for the force is Newton's law of gravitation
\begin{equation}
\vec F_N(\vec r) = - \frac{G_N m M_\odot}{r^2} \frac{\vec r}{r}\ ,
\end{equation}
where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation, $m$ is the mass of Mercury and $M_\odot=1.99\times 10^{30}$ kg is the mass of the Sun.
In addition $r=|\vec r(t)|$ denotes the distance between Sun and Mercury when we assume that the Sun is infinitely heavier than Mercury and located at the center of the coordinate system.
Although this is not exact, it is a pretty good approximation because $m/M_\odot\sim 10^{-8}$.
For later convenience we introduce the Schwarzschild radius of the Sun
\begin{equation}
r_S=\frac{2G_N  M_\odot}{c^2} = 2.95 \ \mbox{km} \ , \label{rsdef}
\end{equation}
where $c=3.00\times 10^8$ m/s denotes the speed of light.
Note that $r_S$ is the characteristic length scale of the gravitational field of the Sun for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and $c$.
The Schwarzschild radius is also an important quantity to characterize black holes; this is however not relevant for the discussion at hand.
With this, Newton's second law reads
\begin{equation}
\ddot{\vec r}      = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\frac{\vec r}{r} \, . \label{eq:newton}
\end{equation}
An attractive force that vanishes at large distances leads in general, depending on the initial conditions, either to bounded or to open orbits.
In the latter case the planet simply disappears from the Sun, since a given force can capture only bodies with small enough momenta.
The students may study those scenarios within their simulations by varying the start velocity while keeping the start location fixed.

@@

To allow for a movement of the perihelion we need to modify the force that led to Eq.~(\ref{eq:newton}).
This modification should be such that it depends on $r$ and it vanishes for $r\to \infty$.
Based on the discussion presented above we may multiply the right hand side of~(\ref{eq:newton}) with a factor
$(1+\alpha(r_S/r))$, where $\alpha$ denotes some dimensionless parameter.
This seems like a natural way to parametrize the additional potential because it adds the smallest possible deviation from an $1/r$ potential expressed relative to the characteristic length $r_S$.
However, besides $r_S$, which characterizes the gravitational potential of the Sun, there is also a characteristic parameter for the dynamics of Mercury:
\begin{equation}
r_a^2 := \frac{\vec L\,^2}{m^2c^2}= \frac{{(\vec r\times \dot{\vec r} \, )}^2}{c^2} \ , \label{a2def}
\end{equation}
where $\vec L$ denotes the angular momentum, which is a constant of motion for central potentials.
It is easily verified that $r_a^2$ carries dimensions of length squared.
We therefore have to add one more term to the potential; in analogy to the one discussed above, we use $\beta (r_a^2/r^2)$.
Combining everything, we use the following ansatz for the modified equation of motion:
\begin{equation}
\ddot{\vec r} = - \frac{c^2}{2}\frac{r_S}{r^2}\left(1+\alpha\frac{r_S}{r}+\beta\frac{r_a^2}{r^2}\right) \, \frac{\vec{r}}{r} \ .
\label{eq:newton_art}
\end{equation}
For the system at hand, $r_a^2$ may be estimated from the parameters of Mercury at its perihelion: The corresponding velocity is
 $\dot{\vec r}(t=0)=\abs{\vec v_M(0)} = 59.0$~km/s (here we already indicate that we will start the simulation at the perihelion) and the closest distance between Sun and Mercury is
$ r_{MS}=\abs{\vec r_{MS}(0)} = 46.0 \cdot 10^6$km~\cite{MercuryFactSheet}. Note that at the perihelion $\vec r$ and $\dot{\vec r}$ are perpendicular to each other.
We thus find that $(r_a^2/ r_{MS}^2) \sim 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})\sim6\cdot 10^{-8}$.
Therefore (for a more detailed discussion about the underlying logic we refer to Sec.~\ref{sec:analysis}) one may expect from both correction terms an effect of similar size. Furthermore, any term of higher order in either $a$ or $r_S$
should be suppressed by additional seven orders of magnitude and thus can be neglected safely.
The parameters $\alpha$ and $\beta$ can be extracted from a fit to data --- this will be done below.
However, since we have only a single number to study, the perihelion shift (see Eq.~(\ref{delT})),
we can only fix either $\alpha$ or $\beta$ (or a linear combination thereof).
The parameters can also be calculated from the underlying theory one uses. The actual values depend on the specific theory;
General relativity gives~\cite{Einstein}
\begin{equation}
	\alpha = 0 \, , \quad \beta = 3 \, . \label{eq:art-prediction}
\end{equation}
Thus we may also use the simulation explained below to calculate the perihelion motion of Mercury from the input
values given in Eq.~(\ref{eq:art-prediction}).

@@

Since the trajectories do not depend on where on the orbit of Mercury the simulation is started, we
use the values at the perihelion with $\abs{\vec r_{MS}(0)} = 46 \cdot 10^6\,\mathrm{km}$ and $\abs{\vec v_M(0)} = 59\,\mathrm{km/s}$\cite{MercuryFactSheet} as initial (for $t=0$) parameters.
The computer does not understand physical units, hence one has to express each variable in an appropriate unit.
Both for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range", e.g.,
by expressing distances in $R_0 = 10^{10}\,\mathrm{m}$ and time intervals in days, $T_0 = 1\,\mathrm{d}$, where d refers to earth days. One Mercury year is given by $T_M=88.0\,T_0$.
With this choice, the initial distance of Mercury to the Sun, the size of the initial velocity of Mercury and the acceleration prefactor (see Eq.~(\ref{eq:newton_art})) become
\begin{eqnarray}
r_{MS}(0)   &=& 4.60 R_0 \, , \quad
v_{M}(0)    = 0.510 \frac{R_0}{T_0} \, ,  \\
a_{MS}(r_{MS}) &=& \frac{c^2}{2}\frac{r_S}{r^2} = 0.990 \frac{R_0}{T_0^2} \frac{1}{{\left(r_{MS}/R_0\right)}^2}\, .
\end{eqnarray}

@@

In \python{}, this reads
\begin{lstlisting}
# Definition of parameters
rM0 = 4.60    # initial radius of Mercury orbit, in units of R0
vM0 = 5.10e-1 # initial orbital speed of Mercury, in units of R0/T0
c_a = 9.90e-1 # base acceleration of Mercury, in units of R0**3/T0**2
TM  = 8.80e+1 # Orbit period of Mercury
rS  = 2.95e-7 # Schwarzschild radius of Sun,in units of R0
ra2 = 8.19e-7 # specific angular momentum, in units of R0**2
\end{lstlisting}
where $\texttt{c\_a} =a_{MS} r_{MS}^2$, the acceleration without the radius factor.
The last two quantities refer to the Schwarzschild radius of the Sun and the parameter $r_a^2$ defined in Eqs. (\ref{rsdef}) and (\ref{a2def}), respectively.

@@

If the starting values are chosen as advised in the previous section, the students should end up with a trajectory as depicted in figure~\ref{fig:MercuryOrbit-a0-small}.
At this point one might ask students to vary $\Delta t$ and observe the effect of this on the orbit [see also figure~\ref{fig:MercuryOrbit-a0-small-dt-large}].
To get the perihelion motion, the additional force term described in Eq.~(\ref{eq:newton_art}) has to be "turned on", e.g., by choosing  $\alpha\neq0$
[figure~\ref{fig:MercuryOrbit-a6-small}].
This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for its impact on the trajectories. In particular
they should discover that values of $\alpha$ of the order of $10^6$ are necessary to get a visible effect.
For a nicely visible perihelion motion it is advisable to not use the correct Mercury values but a more excentric trajectory by choosing, e.g., $r_{MS}(0)=6R_0$ [figure~\ref{fig:MercuryOrbit-a6-big}].
\begin{figure}[htb]
	\centering
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a0T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a0-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/num-err.png}
		\caption{\label{fig:MercuryOrbit-a0-small-dt-large}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T10dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-big}}
	\end{subfigure}
	\captionsetup{singlelinecheck=off}
	\caption[]{\label{fig:MercuryOrbit}
		Different Mercury orbits for $\beta=0$ and
		\begin{itemize}
		\item[(a)] $\alpha=0$ and $\Delta t = \Delta t_0  / 20$;
		\item[(b)] $\alpha=0$ and $\Delta t = \Delta t_0  \times 2$, where the black line represents the orbit of (a) and the red line is the orbit for the larger time steps;
		\item[(c)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$;
		\item[(d)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$ but $r_{MS}(0) = 6 R_0$.
		\end{itemize}
		Here, the time steps are defined by $\Delta t_0 \equiv 2 v_{M}(0) /a_M(0)$ and the images are screenshots of the simulation (with modified colors).
		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "Mercury year" changes.
	}
\end{figure}

@@

\begin{figure}[htb]
	\centering
	\begin{subfigure}[c]{0.49\textwidth}
		\includegraphics[width=\textwidth]{figs/alpha-angle.pdf}
		\caption{\label{fig:angle-alpha}}
	\end{subfigure}
	\begin{subfigure}[c]{0.49\textwidth}
		\includegraphics[width=\textwidth]{figs/beta-angle.pdf}
		\caption{\label{fig:beta-alpha}}
	\end{subfigure}
	\caption{\label{fig:AlphaAngle} Linear relation between $\alpha$ (a), $\beta$ (b) and the perihelion motion $\delta \Theta$ for $\Delta t=2v_M(0)/a_M(0)/200$ and the other parameter set to zero.}
\end{figure}

As an example we  extract the perihelion motion using the parameters provided by General Relativity, namely $\alpha = 0$ and $\beta = 3$.
Considering one result from the numeric simulation, e.g., $\delta\Theta(\alpha = 0, \beta=10^5) = 0.964^{\degree}$, the sought-after angle can be calculated by a linear interpolation as
\begin{equation}
	\delta\Theta (\alpha=0, \beta=3) = \frac{0.964^{\degree}}{10^5} \times 3 = 0.104''
	\, .
\end{equation}
Depending on the knowledge of the students, multiple points and also estimated numerical uncertainties could be used to extract this value by a linear regression.
Here, one has to find a proper range for the interpolation.
Small values for $\alpha$ and $\beta$ usually come with relatively larger numerical errors while for larger values one cannot guarantee that the additional terms are sufficiently small.
For this reason, we suggest to interpolate between zero and $\alpha_{\max}$, $\beta_{\max} \sim 10^5$.
This will be further explained in the next section.
Note that, when comparing the results of the simulation with the experimental result, one Earth year corresponds to $T_E/T_M \approx 365d/88d\approx 4.15$ Mercury years.
Therefore, the perihelion motion per 100 earth years is $0.104''\cdot 415\approx 43.2''$.
This is close to the experimental value of $\delta\Theta_{\mathrm{exp}} = 43.0''$.

@@

\begin{figure}[htb]
	\centering
	\includegraphics[width=.99\textwidth]{figs/angular-variaton.pdf}
	\caption{\label{fcc3}Motion of perihelion $\delta\Theta$ in degrees depending on the number of turns $N_t$ for different time steps $\Delta t$ (color) and different values for $\beta$ (columns).
	Here, $\Delta t_0 \equiv 2 v_M(0)/a_M(0)$.
	Offsets of $\delta \Theta(\alpha=0, \beta=0)$ from zero can be used as an estimated of the magnitude of the error. Oscillations in $\delta \Theta$ indicate too coarse time steps.
}
\end{figure}

@@

This problem can be pointed out to the students by asking them to vary the size of the time-steps and observe the effect on the trajectories.
This should also be done for a wide range of values to show both good and problematic regimes.
It is advantageous to set both $\alpha$ and $\beta$ to zero for this study, since in this case the trajectories should be closed and deviations from the correct case can be spotted easily.
Figure~\ref{fig:MercuryOrbit-a0-small-dt-large} shows the Mercury trajectory for $\Delta t=2 v_M(0) /a_M(0) \times 2$.
One can clearly see the failure of the simulation to reproduce the physical trajectory.
From examples like this it should be apparent that the time steps influence the accuracy of the simulation.
The simulation reproduces the actual trajectory only in the limit $\Delta t \rightarrow 0$. Thus, in any numerical simulation
one always has to identify a proper compromise between numerical accuracy and time spent for the simulation
(clearly for $\Delta t\to 0$ the computing time goes to infinity).

\todo[inline]{Explain fit range. From here until the next section is not complete yet!}
\begin{figure}[htb]
	\centering
	\includegraphics[width=.99\textwidth]{figs/computation-precision.pdf}
	\caption{\label{fcc4}Difference of numerically extracted perihelion angle relative to value computed from General Relativity for different time steps.  Uncertainty bands correspond to equation \ref{eq:num-uncertainty}.
}
\end{figure}

The total numerical uncertainty for the perihelion growth $\Delta \delta \theta (\alpha, \beta)$ at non-zero $\alpha$ and $\beta$ is estimated by the offset of the growth at $\alpha = 0 = \beta$, the oscillation of the growth at $\alpha = 0 = \beta$ as well as the oscillation of the growth at the computed values of $\alpha$ and $\beta$
\begin{eqnarray}\label{eq:num-uncertainty}
	\Delta \delta \Theta (\alpha, \beta) &=& \sqrt{
		\delta \Theta_{std}^2 (0, 0) + \delta \Theta_{avg}^2 (0, 0)+ \delta \Theta_{std}^2 (\alpha, \beta)
	} \, ,
	\\
	\delta \Theta_{avg} (\alpha, \beta)
	&=& 
	\frac{1}{N_{\mathrm{turns}}}
	\sum\limits_{n=1}^{N_{\mathrm{turns}}} \delta \Theta_{n} (\alpha, \beta) \, ,
	\\
	\delta \Theta_{std}^2 (\alpha, \beta)
	&=& 
	\frac{1}{N_{\mathrm{turns}}}
	\sum\limits_{n=1}^{N_{\mathrm{turns}}} \left[
		\delta \Theta_{avg} (\alpha, \beta) - \delta \Theta_{n} (\alpha, \beta)
	\right]^2 \, .
\end{eqnarray}
Thus, the extract numerical value is given by
\begin{equation}
	\delta \Theta (\alpha, \beta) = \delta \Theta_{avg} (\alpha, \beta)  \pm \Delta \delta \Theta (\alpha, \beta)
	\, .
\end{equation}

The prediction coming from General Relativity for zero $\alpha$ is given by
\begin{equation}
	\delta \Theta_{GR} (\alpha=0, \beta) = 
	\frac{1}{4} \frac{r_S^2}{r_a^2} \times \beta
	+\frac{3}{32} \frac{ r_S^4}{ r_a^4} \times \beta ^2
	+O\left(\frac{r_S}{r_a}\right)^6
	\, .
\end{equation}

@@

\todo[inline]{Summary}
